\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.4cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb,bm}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{array}
\usepackage{xcolor}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{url}

\emergencystretch=3em
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\transpose}{^{\mathsf T}}
\newcommand{\code}[1]{\nolinkurl{#1}}

\title{\textbf{Supplementary Material}\\
Reframing preprocessing selection as model-internal calibration in near-infrared spectroscopy}

\author{
Gr{\'e}gory Beurier \quad Robin Reiter \quad Camille No{\^u}s\\[0.35em]
Lauriane Rouan \quad Denis Cornet
}

\date{Draft of May 13, 2026}

\begin{document}
\maketitle

\tableofcontents

\section{Purpose of the Supplement}

This supplement documents the evidence chain behind the main manuscript.  It
is deliberately more operational than the article: local result sources are
listed, claims are mapped to files, deployable and oracle quantities are
separated, and reviewer-sensitive weaknesses are made explicit.  The goal is
to make the AOM study auditable rather than to hide ongoing work behind a
single compressed results table.

The supplement uses local artifacts generated in the repository, including the
refreshed cohort manifest, partial aggregation outputs and the seeds 0/1/2
AOM-PLS robustness run.

For a Talanta submission, the supplement should ultimately become the place
where full dataset tables, paired statistical tests, operator-selection
frequencies, failure cases, leakage audits and software reproducibility
metadata are made explicit.  The present version already establishes that
structure and marks the remaining experiments required before submission.

\section{Notation and Linear-Operator Scope}

Let $\X\in\mathbb{R}^{n\times p}$ be a centered spectral matrix and
$\Y\in\mathbb{R}^{n\times q}$ a centered response matrix.  A strict linear
spectral operator $\A_b\in\mathbb{R}^{p\times p}$ acts as
\begin{equation}
  \X_b = \X\A_b\transpose .
\end{equation}
The operator bank always contains the identity.  This makes standard PLS or
standard Ridge available to the selection procedure.

The strict-linear scope includes smoothing filters, derivative filters,
finite differences and detrending projections.  It does not include
sample-adaptive or fitted transformations such as SNV, MSC, EMSC or ASLS.
Those transformations are useful in NIRS but must be treated as fold-local
pipeline branches.  This distinction is central to the paper because the fast
AOM algebra is valid only for fixed operators inside a calibration fold.

\section{AOM-PLS Derivation}

\subsection{Covariance identity}

For a fixed linear operator $\A_b$, the transformed cross-covariance is
\begin{equation}
  \X_b\transpose \Y =
  (\X\A_b\transpose)\transpose\Y =
  \A_b\X\transpose\Y .
  \label{eq:sup_cov_identity}
\end{equation}
This identity permits candidate operators to be evaluated in covariance space
for SIMPLS and in adjoint form for NIPALS.  It avoids materializing every
transformed matrix $\X\A_b\transpose$ when the operator can be applied directly
to vectors or matrices.

\subsection{Original-space coefficients}

If a PLS component is estimated in transformed space with direction $r_a$, the
corresponding original-space direction is
\begin{equation}
  z_a = \A_b\transpose r_a .
\end{equation}
Collecting $k$ such directions gives $\Z=[z_1,\ldots,z_k]$ and scores
$\mathbf T=\X\Z$.  The final coefficient matrix is
\begin{equation}
  \B = \Z(\mathbf P\transpose\Z)^+\mathbf Q\transpose .
\end{equation}
Thus AOM-PLS remains a single linear calibration model on the original
wavelength grid.

\section{AOM-Ridge Derivation}

For centered $\X_c$ and $\Y_c$, Ridge regression can be solved in the dual.  A
single operator gives
\begin{align}
  \mathbf Z_b &= \X_c\A_b\transpose,\\
  \K_b &= \mathbf Z_b\mathbf Z_b\transpose
       = \X_c\A_b\transpose\A_b\X_c\transpose,\\
  \mathbf C &= (\K_b + \alpha \mathbf I_n)^{-1}\Y_c.
\end{align}
The original-space coefficient matrix is
\begin{equation}
  \bm\beta_b = \A_b\transpose\A_b\X_c\transpose\mathbf C.
\end{equation}

For a superblock or weighted operator mixture,
\begin{equation}
  \K = \sum_{b=1}^{m}s_b^2\X_c\A_b\transpose\A_b\X_c\transpose,
  \qquad
  \mathbf M = \sum_{b=1}^{m}s_b^2\A_b\transpose\A_b,
\end{equation}
and the original-space coefficient matrix is
\begin{equation}
  \bm\beta = \mathbf M\X_c\transpose(\K+\alpha\mathbf I_n)^{-1}\Y_c.
\end{equation}

\section{Algorithmic Sketches}

\subsection{Global AOM-PLS}

\begin{enumerate}
  \item Build a compact operator bank containing identity and a small set of
  smoothing, derivative and detrending operators.
  \item For each candidate number of components and each operator, evaluate
  calibration-fold performance.
  \item Select the operator and component count with the best calibration-fold
  criterion.
  \item Refit on the complete calibration set.
  \item Store selected operator, component count, original-space coefficients,
  centering constants and fold-level diagnostics.
\end{enumerate}

\subsection{AOM-Ridge}

\begin{enumerate}
  \item Build operator-induced kernels
  $\K_b=\X_c\A_b\transpose\A_b\X_c\transpose$ inside the calibration fold.
  \item Select $\alpha$ and operator geometry by nested calibration-fold
  validation.
  \item Solve the dual Ridge system.
  \item Recover original-space coefficients.
  \item Refit on the complete calibration set and evaluate once on the external
  test set.
\end{enumerate}

\section{Operator and Branch Catalogue}

\begin{table}[h]
\centering
\caption{Strict operators and branch preprocessors considered in the AOM study.}
\small
\begin{tabularx}{\linewidth}{p{0.23\linewidth}p{0.35\linewidth}X}
\toprule
Group & Examples & Treatment in the paper \\
\midrule
Identity & $I$ & Always included; recovers the standard model.\\
Smoothing & Savitzky--Golay smoothing & Strict linear operator.\\
Derivatives & SG first and second derivatives, finite difference & Strict linear operator.\\
Detrending & degree-1 and degree-2 trend projections & Strict linear operator when fitted inside folds.\\
Scatter correction & SNV, MSC, EMSC & Fold-local branch preprocessing, not a fixed operator.\\
Baseline correction & ASLS and related smoothers & Fold-local branch preprocessing, not a fixed operator.\\
Orthogonal correction & OSC-style branches & Fold-local branch; not part of the strict AOM algebra.\\
\bottomrule
\end{tabularx}
\end{table}

\section{Empirical Path to the Compact Operator Bank}

The compact operator bank used in the main manuscript is not an arbitrary
small list of preprocessors.  It is the result of a sequence of AOM-specific
experiments whose common conclusion is that, for NIRS calibration with small
external-validation splits, the main risk is not insufficient preprocessing
variety but selection variance over too many near-duplicate operators.  The
compact bank keeps one or two representatives of the main linear spectral
families while avoiding large redundant grids of Savitzky--Golay widths,
Norris--Williams variants and composed operators.

\subsection{Bank genealogy}

\begin{longtable}{L{0.16\linewidth}L{0.09\linewidth}L{0.13\linewidth}L{0.44\linewidth}}
\caption{Operator-bank genealogy used during AOM development.  Candidate
pairs are counted as bank size $\times K_{\max}$ with $K_{\max}=15$, before
fold replication.}\\
\toprule
Bank & Ops & Candidate pairs & Purpose in the study \\
\midrule
\endfirsthead
\toprule
Bank & Ops & Candidate pairs & Purpose in the study \\
\midrule
\endhead
\code{compact} & 9 & 135 &
Minimal strict-linear bank used for the final deployable AOM-PLS
recommendation and for the C++ \code{AOM_lib} implementation.\\
\code{family_pruned} & 15 & 225 &
Pruned version of the production bank, keeping at most two representatives
per preprocessing family.  It tests whether a small but broader family set
beats the compact bank.\\
\code{response_dedup} & 46--47 & 690--705 &
Default-bank reduction by intrinsic-response cosine similarity.  It tests
whether removing near-duplicate operator responses is enough to control
selection variance.\\
\code{default} & 100 & 1500 &
Mirror of the deployed \code{nirs4all} production bank: identity, 32
Savitzky--Golay/Norris--Williams/finite-difference operators, three detrend
projections and 64 SG-by-detrend compositions.\\
\code{deep3} & 116 & 1740 &
Default bank plus third-order strict-linear composition chains.  Used to test
whether deeper composed preprocessors add useful capacity.\\
\code{deep4} & 132 & 1980 &
Default bank plus third- and fourth-order composition chains.  Used as a
stress test for bank-size effects.\\
\code{compact+FCK} & 17 & 255 &
Compact bank augmented with eight fractional convolution kernels.  Used as a
diagnostic extension, not promoted as the default.\\
\bottomrule
\end{longtable}

The final compact bank is exactly:
\begin{equation}
\begin{split}
\mathcal A_{\mathrm{compact}}=\{&
\mathrm{identity},
\mathrm{SG}_{11,2}^{(0)},\mathrm{SG}_{21,3}^{(0)},
\mathrm{SG}_{11,2}^{(1)},\mathrm{SG}_{21,3}^{(1)},\\
&\mathrm{SG}_{11,2}^{(2)},
\mathrm{detrend}_{1},\mathrm{detrend}_{2},
\mathrm{FD}_{1}\}.
\end{split}
\end{equation}
The C++ implementation uses the same ordering:
\code{identity},
\code{sg_smooth_w11_p2},
\code{sg_smooth_w21_p3},
\code{sg_d1_w11_p2},
\code{sg_d1_w21_p3},
\code{sg_d2_w11_p2},
\code{detrend_d1},
\code{detrend_d2},
\code{fd_d1}.

\subsection{Why fewer operators can be better}

For a bank with $B$ operators and $K_{\max}$ component prefixes, a global
AOM selector evaluates approximately $M=B K_{\max}$ operator/component
candidates per selection split, and $F M$ candidates under $F$-fold
cross-validation.  On small calibration sets, validation RMSE estimates have
non-negligible noise.  The empirical minimum over many candidates is therefore
optimistically biased; a crude order-statistic approximation is
\begin{equation}
  \mathbb E[\min_i \widehat R_i]
  \approx
  \mu - \frac{\sigma}{\sqrt{n_{\mathrm{val}}}}\sqrt{2\log M}.
\end{equation}
Thus adding near-duplicate operators can worsen external performance even
when the larger bank strictly contains the smaller one.  In the AOM setting,
moving from \code{compact} to \code{default} increases $M$ from 135 to 1500
at $K_{\max}=15$.  The nominal selection-bias factor
$\sqrt{\log(1500)/\log(135)}$ is about 1.22, which is large enough to matter
for BEER- or BISCUIT-sized splits where the internal validation set contains
only a few samples.  The compact bank was therefore treated as a bias-control
device, not only as a speed optimization.

\subsection{Ablation evidence}

\begin{longtable}{L{0.27\linewidth}L{0.20\linewidth}rrrr}
\caption{AOM-PLS ablations that led to the compact-bank recommendation.
Lower median relative RMSEP is better.  Values are from
\code{bench/AOM_v0/publication/tables/relative_rmsep_per_variant.csv}; full
variant identifiers are preserved in that source table.}\\
\toprule
Variant & Bank / selector & N & Median & Wins & Median fit (s) \\
\midrule
\endfirsthead
\toprule
Variant & Bank / selector & N & Median & Wins & Median fit (s) \\
\midrule
\endhead
nirs4all default & 100-op production holdout & 57 & 0.999 & 29 & 0.96\\
Default NIPALS & 100-op default holdout & 57 & 0.999 & 29 & 8.44\\
Family-pruned & 15-op holdout & 57 & 0.998 & 30 & 0.58\\
Response-dedup & 46--47-op holdout & 57 & 0.998 & 29 & 1.97\\
Compact CV-3 & 9-op CV-3 & 57 & 0.997 & 32 & 0.96\\
Compact CV-5 & 9-op CV-5 & 57 & 0.992 & 38 & 1.23\\
Compact repeated CV-3 & 9-op repeated CV-3 & 57 & 0.984 & 39 & 2.08\\
Response-dedup CV-3 & 46--47-op CV-3 & 57 & 0.988 & 33 & 4.27\\
ASLS compact CV-3 & ASLS + 9-op CV-3 & 57 & 0.979 & 38 & 0.91\\
ASLS compact repeated CV-3 & ASLS + 9-op repeated CV-3 & 57 & 0.975 & 39 & 2.21\\
ASLS compact CV-5 & ASLS + 9-op CV-5 & 57 & 0.960 & 42 & 1.36\\
ASLS family-pruned CV-3 & ASLS + 15-op CV-3 & 57 & 0.964 & 38 & 1.60\\
ASLS response-dedup CV-3 & ASLS + 46--47-op CV-3 & 57 & 0.964 & 37 & 4.57\\
SPXY response-dedup CV-3 & SPXY + 46--47-op CV-3 & 54 & 0.962 & 35 & 29.16\\
\bottomrule
\end{longtable}

These ablations support three practical conclusions.  First, the compact bank
already matches the 100-operator production/default bank at the median while
using roughly one tenth of the operator/component candidates.  Second, the
largest gain comes from replacing deterministic holdout selection with
fold-based selection: compact CV-5 improves from the production/default
median near 0.999 to 0.992 without any ASLS branch, and repeated CV-3 reaches
0.984.  Third, adding the ASLS branch before compact AOM-PLS gives the current
headline variant, but the nearby 15-op and 46--47-op bank results show that
the conclusion is not a one-off artifact of one tiny bank.  The compact bank
is selected as the deployable default because it lies on the accuracy/time
frontier: ASLS + response-dedup with SPXY can be marginally close, but it is
substantially slower and has fewer paired rows.

\subsection{Selected operators inside the compact bank}

\begin{table}[h]
\centering
\caption{Selected first operator in the compact bank on the 57-dataset
cohort.  Counts are taken from the fold-based AOM-PLS result files and show
that the compact bank does not collapse to a single preprocessing choice.}
\small
\begin{tabularx}{\linewidth}{L{0.27\linewidth}Xrr}
\toprule
Operator & Spectral role & AOM CV-5 & ASLS-AOM CV-5 \\
\midrule
\code{identity} & Raw centered spectrum / PLS fallback & 5 & 7\\
\code{sg_smooth_w11_p2} & Short-window SG smoothing & 0 & 1\\
\code{sg_smooth_w21_p3} & Wider SG smoothing & 12 & 17\\
\code{sg_d1_w11_p2} & Short-window first derivative & 9 & 7\\
\code{sg_d1_w21_p3} & Wider first derivative & 8 & 7\\
\code{sg_d2_w11_p2} & Second derivative & 1 & 3\\
\code{detrend_d1} & Linear detrending projection & 8 & 3\\
\code{detrend_d2} & Quadratic detrending projection & 8 & 4\\
\code{fd_d1} & Finite difference first derivative & 6 & 8\\
\bottomrule
\end{tabularx}
\end{table}

The selected-operator distribution is important for interpretation.  The
identity remains selected on several datasets, which is expected when raw
spectra already carry the analyte signal.  The most frequent non-identity
operator is the wider SG smoother, followed by derivative and detrending
operators.  This supports the design choice that the compact bank should cover
distinct spectral mechanisms rather than dense parameter grids within one
mechanism.

\subsection{Extensions that were not promoted}

Several extensions were explored and intentionally not moved into the compact
default.

\begin{itemize}
  \item The 100-operator default bank and the deep3/deep4 banks did not improve
  the median external RMSEP enough to justify the added selection variance and
  runtime.  Deep banks mostly reproduced the default-bank choice on the tested
  splits.
  \item OSC and EMSC branches were not promoted as defaults.  Existing
  AOM\_v0 summaries show OSC variants hurting at the default
  component setting and EMSC degree 1--2 being numerically unstable on small
  splits.
  \item SNV can help as a branch, but SNV+ASLS did not improve over ASLS alone
  in the 57-dataset summary.  SNV, MSC, EMSC and ASLS remain branch
  preprocessors, not strict fixed operators in the AOM algebra.
  \item Fractional convolution kernel operators were implemented as diagnostic
  bank extensions (\code{compact+FCK}: 9 compact operators plus eight FCK
  kernels).  They remain useful evidence that fractional-derivative
  preprocessors were considered, but they are not part of the current compact
  recommendation until they pass the same paired, multi-seed and no-harm gates
  as the compact AOM-PLS variants.
\end{itemize}

\section{Benchmark Protocol Ledger}

\begin{longtable}{L{0.28\linewidth}L{0.29\linewidth}L{0.35\linewidth}}
\caption{Local evidence map used to prepare the manuscript.}\\
\toprule
Artifact & Content & Use in the paper \\
\midrule
\endfirsthead
\toprule
Artifact & Content & Use in the paper \\
\midrule
\endhead
\code{bench/AOM_v0/Summary.md} &
57-dataset AOM-PLS summary; compact ASLS+AOM CV-5 champion. &
Primary AOM-PLS headline source.\\
\code{bench/AOM_v0/publication/tables/relative_rmsep_per_variant.csv} &
Relative RMSEP and fit-time summaries for AOM-PLS variants. &
Source for median RMSEP/PLS, wins and fit time.\\
\code{bench/AOM_v0/docs/AOMPLS_MATH_SPEC.md} &
AOM-PLS operator algebra, NIPALS/SIMPLS conventions. &
Method verification and notation.\\
\code{bench/AOM_v0/docs/AOMPLS_VALIDATION.md} &
Equivalence tests: identity, fixed operator, NumPy/Torch parity. &
Implementation validation.\\
\code{bench/AOM_v0/Ridge/docs/AOM_RIDGE_MATH_SPEC.md} &
Dual AOM-Ridge derivation and branch warnings. &
Ridge method section.\\
\code{bench/AOM_v0/Ridge/publication/tables/table_per_method_summary.tex} &
Deployable AOM-Ridge variants on 52 datasets. &
Conservative deployable Ridge claim.\\
\code{bench/AOM_v0/Ridge/publication/tables/table_summary.tex} &
Oracle envelope against reference baselines. &
Upper-bound Ridge claim only.\\
\code{bench/AOM_v0/Ridge/docs/HEADLINE_SPXY3_NESTED_AUDIT.md} &
Nested-selection audit for headline Ridge variants. &
Leakage-risk discussion.\\
\code{bench/tabpfn_paper/article/main.tex} &
Reference benchmark protocol, preprocessing search budgets and splits. &
Baseline protocol context; model name not emphasized in the article.\\
\code{bench/benchmark_master_results.csv} &
Unified registry of many variants and maturity levels. &
Exploration registry, not a direct source of all headline statistics.\\
\bottomrule
\end{longtable}

\subsection{Cohort manifest}

The frozen review manifest is stored at
\code{paper_aom/review/cohort_manifest.csv}.  It contains 78 source rows:
61 regression rows and 17 classification rows.  For primary paired analyses,
the effective cohort is 60 regression datasets plus 16 classification
datasets: the QUARTZ regression row is retained for absolute-error and failure
tables but excluded from ratio summaries because the reference denominator is
near zero, and one classification row is excluded because source files are
missing.  The companion denominator document
\code{paper_aom/review/cohort_manifest.md} records the gates for TabPFN-allowed
subsets and AOM-Ridge-global allowability.

\subsection{Software artefacts}

Table~\ref{tab:sup_software} mirrors the software-status table used in the
main manuscript.  It separates production \code{nirs4all} code, the new
AOM-lib-backed wrapper, parity-verified C++/Python \code{AOM_lib} components,
code-ready bindings that were not locally executed in this validation
environment, and AOM-Ridge research code.

\begin{table}[h]
\centering
\caption{Software artefacts distributed with the AOM study.}
\label{tab:sup_software}
\small
\input{tables/table_software.tex}
\end{table}

\subsection{Multi-seed regression robustness}

The refreshed AOM-PLS robustness run is
\code{bench/scenarios/runs/paper_aom_aompls_seeds012/results.csv}.  It covers
55 datasets, 9 variants and seeds 0/1/2, with 1486 CSV lines at validation
time.  For the headline comparison
\code{ASLS-AOM-compact-cv5-numpy} versus \code{PLS-standard-numpy}, the 53
paired datasets give a median RMSEP/PLS ratio of 0.9623 and 37/53 wins after
averaging RMSEP over seeds.  This is close to the earlier single-seed median
0.960, but the win denominator changes from 57 to 53 under the refreshed
paired multi-seed run.

\begin{table}[h]
\centering
\caption{Seed-stability summary emitted by
\code{paper_aom/review/final_stats.md}.  Median and IQR are the raw RMSEP
dispersion summaries across available seeds, so they are used here as
stability diagnostics rather than cross-trait effect sizes.}
\small
\begin{tabularx}{\linewidth}{Xrrrr}
\toprule
Variant & Datasets & Seeds & Median & IQR \\
\midrule
ASLS-AOM-compact-cv5-numpy & 59 & 3 & 0.4515 & 3.3055\\
AOM-compact-cv5-numpy & 59 & 3 & 0.6327 & 3.3118\\
AOMRidge-Blender-headline-spxy3 & 20 & 3 & 0.3300 & 2.2792\\
AOMRidge-Local-compact-knn50 & 20 & 3 & 0.5549 & 2.5818\\
AOMRidge-AutoSelect-headline-spxy3 & 20 & 3 & 0.3455 & 2.2845\\
AOMRidge-global-compact-none & 20 & 3 & 0.4153 & 2.7123\\
Ridge-tabpfn-hpo-60trials & 7 & 3 & 0.4257 & 3.2861\\
PLS-default-cv5 & 7 & 3 & 0.5713 & 3.4682\\
PLS-tabpfn-hpo-25trials & 7 & 3 & 0.3535 & 4.1037\\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Parity validation}

The dedicated \code{AOM_lib} implementation was validated against the
reference Python AOM\_v0 calculations on BEER, CORN and ALPINE under three
split/selection regimes: kfold5, kfold5+oneSE and spxy5.  The local validation
commands \code{cpp/build/test_operators},
\code{cpp/build/test_parity_kfold cpp/tests/reference} and
\code{python/tests/test_parity.py} passed.  Across the nine parity cases, the
largest reported coefficient difference was below $10^{-8}$, with prediction
and RMSE-curve differences also below the tolerance used by the parity tests.

\section{Claim Ledger}

\begin{longtable}{L{0.29\linewidth}L{0.19\linewidth}L{0.13\linewidth}L{0.23\linewidth}}
\caption{Main claims and their current evidence status.}\\
\toprule
Claim & Status & Dataset count & Source / caveat \\
\midrule
\endfirsthead
\toprule
Claim & Status & Dataset count & Source / caveat \\
\midrule
\endhead
ASLS + compact AOM-PLS CV-5 improves over PLS with median RMSEP/PLS 0.9623. &
Multi-seed AOM-PLS robustness result. & 53 &
\code{paper_aom_aompls_seeds012/results.csv}; paired after averaging over seeds.\\
AOM-PLS compact CV-5 wins against PLS on 37 datasets. &
Multi-seed AOM-PLS robustness result. & 53 &
Same source; replaces the earlier single-seed 57-dataset headline.\\
AOM-PLS compact CV-5 median fit time is approximately 1.65 s. &
Measured in refreshed aggregation. & 339 &
\code{paper_aom/review/final_stats.md}; hardware- and partial-run dependent.\\
AOM-Ridge Blender has median RMSEP/Ridge-HPO ratio 0.905 in the refreshed overlap. &
Partial regenerated evidence. & 4 &
\code{paper_aom/review/final_stats.md}; denominator limited by incomplete HPO overlap.\\
AOM-Ridge Blender wins 4/4 against Ridge-HPO in the refreshed overlap. &
Partial regenerated evidence. & 4 &
Full-cohort Blender/AutoSelect multi-seed promotion is pending.\\
AOM-Ridge oracle improves over tuned Ridge by median 4.73\%. &
Oracle envelope only. & 52 &
\code{table_summary.tex}; not a deployable method claim.\\
AOM-Ridge oracle wins 45/52 against tuned Ridge. &
Oracle envelope only. & 52 &
Use only as headroom / upper-bound evidence.\\
Large preprocessing searches evaluate thousands of fold-level candidates. &
Reference protocol fact. & per dataset &
PLS/Ridge approximately 5400, CNN approximately 2700, CatBoost approximately 90.\\
\bottomrule
\end{longtable}

\section{AOM-PLS Variant Summary}

\begin{table}[h]
\centering
\caption{Historical single-seed AOM-PLS variants from the existing
57-dataset summary.  The refreshed multi-seed headline is reported in the
robustness subsection above.}
\small
\begin{tabularx}{\linewidth}{Xrrrr}
\toprule
Variant & Datasets & Median RMSEP/PLS & Wins & Median fit (s) \\
\midrule
ASLS-AOM compact CV-5 & 57 & 0.960 & 42 & 1.36\\
ASLS-AOM response-dedup CV-3 & 57 & 0.964 & 37 & 4.57\\
ASLS-AOM family-pruned CV-3 & 57 & 0.964 & 38 & 1.60\\
ASLS-AOM compact repeated CV-3 & 57 & 0.975 & 39 & 2.21\\
ASLS-AOM compact CV-3 & 57 & 0.979 & 38 & 0.91\\
\bottomrule
\end{tabularx}
\end{table}

\section{AOM-Ridge Result Separation}

\begin{table}[h]
\centering
\caption{Single-seed AOM-Ridge deployable variants retained as background
evidence before full multi-seed promotion; full local identifiers are in the
source table.}
\small
\begin{tabularx}{\linewidth}{Xrrrr}
\toprule
Variant & Wins / N & Win-rate (\%) & Mean rank & Median $\Delta$ vs Ridge (\%) \\
\midrule
Blender headline SPXY3 & 35/52 & 67.3 & 3.56 & -2.22\\
AutoSelect headline SPXY3 & 27/52 & 51.9 & 4.61 & -0.61\\
Global compact none & 29/52 & 55.8 & 4.79 & -0.63\\
\bottomrule
\end{tabularx}
\end{table}

\begin{table}[h]
\centering
\caption{AOM-Ridge oracle envelope.  These values describe headroom and must
not be worded as deployable practitioner performance.}
\small
\begin{tabularx}{\linewidth}{Xrrrr}
\toprule
Baseline & Median $\Delta$ (\%) & Mean $\Delta$ capped (\%) & Win-rate (\%) & Wins / N\\
\midrule
Ridge & -4.73 & -12.37 & 86.5 & 45/52\\
PLS & -7.73 & -14.66 & 90.4 & 47/52\\
CNN & -12.56 & -16.08 & 74.5 & 35/47\\
CatBoost & -13.27 & -19.77 & 73.1 & 38/52\\
\bottomrule
\end{tabularx}
\end{table}

\section{Paired Statistical Evidence}

\begin{table}[h]
\centering
\caption{Current paired statistical evidence for the two primary deployable
claims.  These values were computed from existing result files and should be
regenerated after the final Talanta cohort manifest is frozen.}
\small
\input{tables/table_paired_stats.tex}
\end{table}

\section{Leakage and Audit Controls}

\begin{table}[h]
\centering
\caption{Leakage controls that must be preserved in any final rerun.}
\small
\begin{tabularx}{\linewidth}{p{0.28\linewidth}X}
\toprule
Control & Requirement \\
\midrule
External test set & Never used for operator, branch, component or $\alpha$ selection.\\
Original splits & Preserved when supplied by the dataset source.\\
Generated splits & Deterministic SPXY or stratified variants, with split type logged.\\
Branch preprocessing & Fitted inside each calibration fold only.\\
Operator selection & Performed using calibration-fold validation only.\\
Final refit & Refit selected configuration on the full calibration set, then evaluate once on test.\\
Oracle tables & Clearly labeled as retrospective upper bounds.\\
Skipped datasets & Logged with inclusion/exclusion reason.\\
\bottomrule
\end{tabularx}
\end{table}

\section{Additional Experiments Required for a Journal Submission}

\begin{longtable}{L{0.13\linewidth}L{0.35\linewidth}L{0.44\linewidth}}
\caption{Experiment backlog.}\\
\toprule
Priority & Experiment & Purpose \\
\midrule
\endfirsthead
\toprule
Priority & Experiment & Purpose \\
\midrule
\endhead
P0 & Freeze one regression cohort manifest. & Remove denominator drift across 52/57/58/61 dataset statements.\\
P0 & Regenerate headline tables from one manifest. & Ensure PLS, Ridge, CatBoost, CNN, AOM-PLS and AOM-Ridge are compared on a common subset.\\
P0 & Separate deployable and oracle tables. & Prevent reviewers from interpreting oracle envelopes as methods.\\
P0 & Add exclusion table. & Document QUARTZ-style degenerate denominators, missing CNN rows and compute skips.\\
P1 & Full multi-seed rerun. & Quantify selection variance and win-rate uncertainty.\\
P1 & Paired statistical tests. & Add Wilcoxon/Friedman/Nemenyi or equivalent tests on the common subset.\\
P1 & Strict-linear vs branch ablation. & Separate gains from AOM itself and gains from ASLS/SNV/MSC branches.\\
P1 & Operator-frequency table. & Show which spectral operations are actually selected.\\
P1 & Runtime scaling study. & Quantify cost versus $n$, $p$, bank size and branch count.\\
P1 & Prediction artifact export. & Enable residual plots, bias, RPD/RPIQ, and independent audit.\\
P1 & Software release validation. & Maintain C++/Python parity validation and complete local execution for the R, MATLAB, Julia and JavaScript/WASM bindings before release promotion.\\
\bottomrule
\end{longtable}

\section{Reproducibility Checklist}

\begin{enumerate}
  \item Record repository commit hash, Python version and package versions.
  \item Store the cohort manifest used for every table.
  \item Store split indices for every dataset.
  \item Store fold predictions and final test predictions.
  \item Store selected operators, component counts, $\alpha$ values and branch
  preprocessing choices.
  \item Store runtime and hardware metadata.
  \item Regenerate tables from scripts rather than manually editing manuscript
  tables.
\end{enumerate}

\section{Recommended Positioning}

The safest wording for the current draft is: AOM is a general linear-operator
framework; AOM-PLS has the strongest deployable full-cohort evidence; AOM-Ridge
is a mathematically consistent and promising extension with deployable evidence
and a stronger oracle envelope, but requires full multi-seed promotion before
being described as a production default.  This position is scientifically
stronger than claiming universal benchmark dominance.

\nocite{geladi1986partial,wold2001pls,dejong1993simpls,hoerl1970ridge,rinnan2009review,savitzky1964smoothing,barnes1989standard,geladi1985linearization}

\bibliographystyle{unsrtnat}
\bibliography{references}

\end{document}
